\chapter{Results}
\label{chap:results}

\section{Overview}

The final results support one dominant practical conclusion: LLM checkpoint loading should be treated as an adaptive heuristic problem. At the Rust layer, smaller eager loads favour a simpler synchronous path, while larger checkpoints increasingly reward the final multi-worker \texttt{io\_uring} backend. The exact crossover depends on format, but the broad rule is stable. Python and vLLM then show how much of that backend advantage survives once higher-layer overheads are introduced.

This chapter therefore presents the evidence in three layers. First, the Rust harness isolates checkpoint loading itself and provides the cleanest backend-attribution results. Second, the Python benchmarks show the cost of crossing the binding and tensor-materialisation boundary. Third, the vLLM matrix evaluates the same checkpoint-loading choices inside a serving stack where model initialisation, TTFT, and decode have different sensitivities to loader choice.

\section{Rust-Layer Backend Results}

The Rust harness is the authoritative backend-attribution layer because it measures checkpoint loading directly, without Python tensor conversion or serving-engine initialisation costs. The full cold-cache matrix is stored in \texttt{results/h100/profile/full\_cold\_matrix.tsv}.

\subsection{SafeTensors}

Table~\ref{tab:rust-safetensors} summarises the explicit backend comparison for \texttt{SafeTensors}. Two regimes are immediately visible.

\begin{table}[h]
\caption{Rust cold-cache \texttt{SafeTensors} results (ms).}
\label{tab:rust-safetensors}
\centering
\begin{tabular}{lrrrr}
\toprule
Model & Sync & Async & Default & io\_uring \\
\midrule
Qwen3-0.6B & 555.28 & 1416.19 & 333.59 & 1159.24 \\
SmolLM3-3B & 1331.79 & 5802.05 & 3722.35 & 3812.35 \\
Qwen3-4B & 1820.85 & 5039.21 & 3083.18 & 2967.42 \\
Qwen3-8B & 3880.40 & 7074.40 & 3844.35 & 3172.37 \\
Qwen3-14B & 7030.25 & 13489.00 & 6138.70 & 5977.07 \\
Qwen3-32B & 15454.45 & 28763.46 & 12119.50 & 12089.05 \\
\bottomrule
\end{tabular}
\end{table}

\begin{figure}[h]
\centering
\begin{tikzpicture}
\begin{axis}[
width=0.95\textwidth,
height=7cm,
ybar,
bar width=6pt,
ymode=log,
ylabel={Cold load time (ms)},
xlabel={Model},
symbolic x coords={0.6B,3B,4B,8B,14B,32B},
xtick=data,
legend style={at={(0.5,1.18)},anchor=south,legend columns=4},
ymajorgrids=true,
]
\addplot coordinates {(0.6B,555.28) (3B,1331.79) (4B,1820.85) (8B,3880.40) (14B,7030.25) (32B,15454.45)};
\addplot coordinates {(0.6B,1416.19) (3B,5802.05) (4B,5039.21) (8B,7074.40) (14B,13489.00) (32B,28763.46)};
\addplot coordinates {(0.6B,333.59) (3B,3722.35) (4B,3083.18) (8B,3844.35) (14B,6138.70) (32B,12119.50)};
\addplot coordinates {(0.6B,1159.24) (3B,3812.35) (4B,2967.42) (8B,3172.37) (14B,5977.07) (32B,12089.05)};
\legend{sync,async,default,io\_uring}
\end{axis}
\end{tikzpicture}
\caption{Rust cold-cache \texttt{SafeTensors} crossover behaviour. The main regime change is from \texttt{sync}-favoured smaller models to \texttt{io\_uring}-favoured larger models.}
\label{fig:rust-safetensors-crossover}
\end{figure}

For smaller eager whole-file loads, synchronous loading remains highly effective. On SmolLM3-3B and Qwen3-4B, \texttt{sync} is the strongest explicit backend. The ranking flips on larger multi-shard checkpoints. On Qwen3-8B, Qwen3-14B, and Qwen3-32B, the multi-worker \texttt{io\_uring} backend is the fastest explicit backend. The resulting rule is the clearest high-level heuristic in the project: \texttt{sync} for smaller \texttt{SafeTensors} checkpoints and \texttt{io\_uring} for larger ones.

The \texttt{async} backend is consistently the weakest choice for \texttt{SafeTensors}. This matches the workload structure: whole-file shard loading benefits from strong bulk parallelism, not from scheduler-heavy fine-grained async orchestration. Repeated five-run anchor measurements reinforce the same split. For the anchor set, the mean best backend is \texttt{sync} for Qwen3-0.6B and \texttt{io\_uring} for Qwen3-8B and Qwen3-32B.

\subsection{ServerlessLLM}

The \texttt{ServerlessLLM} layout produces a different ordering, shown in Table~\ref{tab:rust-serverlessllm}.

\begin{table}[h]
\caption{Rust cold-cache \texttt{ServerlessLLM} results (ms).}
\label{tab:rust-serverlessllm}
\centering
\begin{tabular}{lrrrr}
\toprule
Model & Sync & Async & Default & io\_uring \\
\midrule
Qwen3-0.6B & 1645.06 & 1471.34 & 1432.25 & 1387.57 \\
SmolLM3-3B & 3941.88 & 2391.24 & 3182.28 & 3091.49 \\
Qwen3-4B & 5137.72 & 3504.25 & 3551.82 & 3727.14 \\
Qwen3-8B & 11048.40 & 9463.37 & 7589.47 & 9043.51 \\
Qwen3-14B & 18661.25 & 16139.08 & 13422.94 & 13257.23 \\
Qwen3-32B & 39822.09 & 29180.89 & 27659.78 & 29779.86 \\
\bottomrule
\end{tabular}
\end{table}

\begin{figure}[h]
\centering
\begin{tikzpicture}
\begin{axis}[
width=0.95\textwidth,
height=7cm,
ybar,
bar width=6pt,
ymode=log,
ylabel={Cold load time (ms)},
xlabel={Model},
symbolic x coords={0.6B,3B,4B,8B,14B,32B},
xtick=data,
legend style={at={(0.5,1.18)},anchor=south,legend columns=4},
ymajorgrids=true,
]
\addplot coordinates {(0.6B,1645.06) (3B,3941.88) (4B,5137.72) (8B,11048.40) (14B,18661.25) (32B,39822.09)};
\addplot coordinates {(0.6B,1471.34) (3B,2391.24) (4B,3504.25) (8B,9463.37) (14B,16139.08) (32B,29180.89)};
\addplot coordinates {(0.6B,1432.25) (3B,3182.28) (4B,3551.82) (8B,7589.47) (14B,13422.94) (32B,27659.78)};
\addplot coordinates {(0.6B,1387.57) (3B,3091.49) (4B,3727.14) (8B,9043.51) (14B,13257.23) (32B,29779.86)};
\legend{sync,async,default,io\_uring}
\end{axis}
\end{tikzpicture}
\caption{Rust cold-cache \texttt{ServerlessLLM} crossover behaviour. The range-heavy layout shifts the strong contenders from \texttt{sync} toward \texttt{async} and \texttt{io\_uring}.}
\label{fig:rust-serverless-crossover}
\end{figure}

Here, \texttt{sync} is never the best backend. For smaller and mid-sized range-heavy workloads, \texttt{async} is often strong because grouped and coalesced range reads fit its execution model well. For larger cases, \texttt{io\_uring} becomes competitive or best. For example, Qwen3-14B is fastest with \texttt{io\_uring}, while SmolLM3-3B and Qwen3-4B favour \texttt{async}. The partitioned layout therefore changes the ranking and makes a fixed backend recommendation less defensible.

\subsection{Selector Behaviour}

The final selector is not perfect on every single Rust matrix row, but it is materially better than the earlier cost-model-based selector. On the Rust matrix, \texttt{default} is exactly best for Qwen3-8B and Qwen3-32B in \texttt{ServerlessLLM}, and very close to best for the large \texttt{SafeTensors} cases. The remaining misses are concentrated around the medium-size crossover region, especially SmolLM3-3B and Qwen3-4B. The important point is that the selector now follows the correct \emph{regimes}: smaller \texttt{SafeTensors} loads lean toward \texttt{sync}; larger \texttt{SafeTensors} loads lean toward \texttt{io\_uring}; \texttt{ServerlessLLM} switches between \texttt{async} and \texttt{io\_uring} depending on workload size and structure.

\section{Python Binding Results}

Python benchmarks were collected on Qwen/Qwen3-8B to quantify user-facing overhead and to evaluate whether binding optimisation and selector redesign improved practical performance.

\subsection{Binding Overhead Before and After Optimisation}

The most important Python result is that binding engineering mattered. Replacing per-call runtime setup with a shared Tokio runtime and reducing repeated Torch conversion work materially improved the public \texttt{default} API.

\begin{table}[h]
\caption{Python eager-load times for Qwen3-8B before and after optimisation (seconds).}
\label{tab:python-optimisation}
\centering
\begin{tabular}{lrrr}
\toprule
Case & Early default & Post-binding optimisation & Post-selector redesign \\
\midrule
SafeTensors default & 15.8831 & 14.5451 & 11.5498 \\
ServerlessLLM default & 22.4547 & 20.5008 & 18.4518 \\
\bottomrule
\end{tabular}
\end{table}

\begin{figure}[h]
\centering
\begin{tikzpicture}
\begin{axis}[
width=0.85\textwidth,
height=6.5cm,
ybar,
bar width=12pt,
ylabel={Time (s)},
symbolic x coords={early,binding-opt,selector-redesign},
xtick=data,
legend style={at={(0.5,1.12)},anchor=south,legend columns=2},
ymajorgrids=true,
]
\addplot coordinates {(early,15.8831) (binding-opt,14.5451) (selector-redesign,11.5498)};
\addplot coordinates {(early,22.4547) (binding-opt,20.5008) (selector-redesign,18.4518)};
\legend{SafeTensors default,ServerlessLLM default}
\end{axis}
\end{tikzpicture}
\caption{Python \texttt{default} eager-load times on Qwen3-8B before and after binding and selector improvements.}
\label{fig:python-default-improvement}
\end{figure}

For \texttt{SafeTensors} on Qwen3-8B, the \texttt{default} path improves from 15.8831~s to 11.5498~s, a 27.3\% reduction relative to the early Python baseline and a 20.6\% reduction relative to the post-binding baseline. For \texttt{ServerlessLLM}, the same sequence improves \texttt{default} from 22.4547~s to 18.4518~s, an overall reduction of 17.8\% and a 10.0\% reduction relative to the post-binding baseline.

\subsection{Python Layer Interpretation}

The Python numbers are much larger than the Rust-only numbers because eager loading at the binding layer includes substantial tensor materialisation and Python object construction. This is especially visible in the \texttt{SafeTensors} benchmark, where native \texttt{safetensors} loading remains much faster than the fully materialised \texttt{tensor\_store} eager paths. The key conclusion is that Python is not a transparent pass-through layer. Backend improvements matter, but they are filtered through the cost of converting checkpoint contents into Torch tensors and Python dictionaries.

\section{vLLM Results}

The final vLLM matrix completed successfully across all six models, all benchmark loaders, warm and cold cache conditions, and all three benchmark kinds. The completed summary is stored in \texttt{results/h100/vllm/full\_matrix.tsv}. The final matrix contains 672 successful rows and no failures.

\subsection{Cold Load-Only Results}

For cold \texttt{load\_only}, the best loader varies with model size. Table~\ref{tab:vllm-load-only} shows the strongest loader in the final cold-load-only configuration.

\begin{table}[h]
\caption{Best cold \texttt{load\_only} vLLM loader by model.}
\label{tab:vllm-load-only}
\centering
\begin{tabular}{lrl}
\toprule
Model & Best init time (ms) & Best loader \\
\midrule
Qwen3-0.6B & 20417.42 & ts\_safetensors\_sync \\
SmolLM3-3B & 27333.96 & ts\_safetensors\_default \\
Qwen3-4B & 28475.27 & ts\_safetensors\_default \\
Qwen3-8B & 35555.79 & ts\_safetensors\_default \\
Qwen3-14B & 47751.38 & ts\_safetensors\_io\_uring \\
Qwen3-32B & 88164.00 & ts\_safetensors\_io\_uring \\
\bottomrule
\end{tabular}
\end{table}

\begin{figure}[h]
\centering
\begin{tikzpicture}
\begin{axis}[
width=0.95\textwidth,
height=7cm,
ybar,
bar width=7pt,
ymode=log,
ylabel={Cold init time (ms)},
xlabel={Model},
symbolic x coords={0.6B,3B,4B,8B,14B,32B},
xtick=data,
legend style={at={(0.5,1.18)},anchor=south,legend columns=4},
ymajorgrids=true,
]
\addplot coordinates {(0.6B,20577.94) (3B,27333.96) (4B,28475.27) (8B,35555.79) (14B,47792.01) (32B,88199.98)};
\addplot coordinates {(0.6B,20417.42) (3B,27767.25) (4B,28478.77) (8B,36524.10) (14B,49435.10) (32B,92421.39)};
\addplot coordinates {(0.6B,20867.14) (3B,29992.67) (4B,29727.16) (8B,35889.70) (14B,47751.38) (32B,88164.00)};
\addplot coordinates {(0.6B,21191.06) (3B,29364.71) (4B,31792.30) (8B,42343.08) (14B,60669.99) (32B,111717.78)};
\legend{ST default,ST sync,ST io\_uring,SLLM default}
\end{axis}
\end{tikzpicture}
\caption{Cold vLLM \texttt{load\_only} times for the strongest loader families. SafeTensors loaders dominate, and \texttt{io\_uring} becomes best at the large end of the ladder.}
\label{fig:vllm-load-only}
\end{figure}

Two observations stand out. First, the best vLLM loader is usually from the \texttt{SafeTensors} family rather than the \texttt{ServerlessLLM} family. Second, the explicit \texttt{io\_uring} path becomes valuable again at the larger end of the ladder. It is the best \texttt{load\_only} choice for Qwen3-14B and Qwen3-32B, albeit by a narrow margin over the adaptive \texttt{default} path (0.09\% on 14B and 0.04\% on 32B). For smaller models, the best loader is usually \texttt{sync} or \texttt{default}, which mirrors the core Rust heuristic story.

\subsection{TTFT and Steady-State Decode}

The backend effect is strongest for initialisation and TTFT, and much weaker for steady-state decode. Table~\ref{tab:vllm-ttft} shows the best cold TTFT result per model and the margin to the second-best loader.

\begin{table}[h]
\caption{Best cold TTFT result by model, with margin to second-best loader.}
\label{tab:vllm-ttft}
\centering
\begin{tabular}{lrrl}
\toprule
Model & Best TTFT (ms) & Margin & Best loader \\
\midrule
Qwen3-0.6B & 20272.60 & 0.21\% & ts\_safetensors\_sync \\
SmolLM3-3B & 27463.41 & 1.22\% & ts\_safetensors\_sync \\
Qwen3-4B & 28529.77 & 0.26\% & ts\_safetensors\_default \\
Qwen3-8B & 35099.84 & 0.80\% & ts\_safetensors\_default \\
Qwen3-14B & 48201.50 & 1.59\% & ts\_safetensors\_default \\
Qwen3-32B & 88550.77 & 0.83\% & ts\_safetensors\_default \\
\bottomrule
\end{tabular}
\end{table}

For cold TTFT, \texttt{ts\_safetensors\_default} is strongest on four of the six models, while \texttt{ts\_safetensors\_sync} is strongest on the two smallest models. The margins are small but consistent, which is exactly what one would expect once model loading is only one component of the overall TTFT path.

The steady-state decode results tell a different story. Table~\ref{tab:vllm-decode} summarises the best cold decode result per model. The differences are tiny: the best-vs-second-best margin never exceeds 0.67\% and is usually much smaller.

\begin{table}[h]
\caption{Best cold steady-state decode result by model.}
\label{tab:vllm-decode}
\centering
\begin{tabular}{lrrl}
\toprule
Model & Best decode (ms) & Margin & Best loader \\
\midrule
Qwen3-0.6B & 17.21 & 0.67\% & native \\
SmolLM3-3B & 35.91 & 0.39\% & ts\_serverlessllm\_default \\
Qwen3-4B & 39.02 & 0.06\% & ts\_safetensors\_io\_uring \\
Qwen3-8B & 58.10 & 0.05\% & ts\_serverlessllm\_io\_uring \\
Qwen3-14B & 94.21 & 0.19\% & ts\_safetensors\_default \\
Qwen3-32B & 194.20 & 0.02\% & native \\
\bottomrule
\end{tabular}
\end{table}

This confirms that backend choice is primarily a cold-start and model-initialisation problem rather than a decode-throughput problem. Once the model is resident, GPU compute dominates and the loader's influence becomes marginal.

\subsection{Large-Model Feasibility}

The 14B and 32B models initially failed in vLLM. These were not backend failures. The raw logs showed that the benchmark harness was too conservative in its vLLM memory budget, leaving insufficient KV-cache allocation headroom. After increasing the vLLM memory budget and reducing \texttt{max\_model\_len}, all previously failing rows succeeded. This matters for interpretation: the initial failures were a serving-harness configuration issue, not evidence that the checkpoint backends could not support the large models.

\section{Negative and Turnaround Results}

The most important negative result in the project was the early single-ring \texttt{io\_uring} prototype. On the H100 system, that implementation was much slower than \texttt{sync} and showed poor CPU utilisation. Profiling pointed to a design problem rather than a fundamental backend limitation: one ring and one issuer thread left too much of the machine idle.

The project changed direction after that result. Once the \texttt{io\_uring} backend was redesigned around multiple worker-owned rings, its ranking changed significantly. It became competitive or best on larger \texttt{SafeTensors} and \texttt{ServerlessLLM} workloads. The final result is therefore not ``io\_uring always wins'', nor ``io\_uring is a dead end''. The correct conclusion is that backend family and backend architecture both matter.

\section{Summary}

Three final findings emerge from the completed result set.

\begin{enumerate}
\item At the Rust layer, backend ranking depends strongly on workload structure. \texttt{SafeTensors} favours \texttt{sync} on smaller cases and \texttt{io\_uring} on larger cases. \texttt{ServerlessLLM} favours \texttt{async} or \texttt{io\_uring}, while \texttt{sync} is rarely best.
\item At the Python layer, binding overhead is large enough to matter, and binding-specific optimisation produced meaningful practical gains.
\item At the vLLM layer, backend choice affects initialisation and TTFT far more than steady-state decode, and the best loader depends on model size.
\end{enumerate}

Taken together, the results support an adaptive, format-aware checkpoint-loading heuristic. The dominant practical rule is sync for smaller models and io\_uring for larger ones, with additional format-specific adjustments for range-heavy layouts.
