\appendix
\chapter{Backend Interface Specifications}
\label{app:backend_api}

\section{Rust Backend API}

The core Rust backend interface for synchronous loading is defined in the Python bindings at \texttt{safetensors.rs} and \texttt{serverlessllm.rs}, within the \texttt{bindings/python/rust\_ext/src/api/} package directory.

\subsection{SafeTensors Synchronous Loading}

The synchronous loading function validates the path, loads the checkpoint using the Rust \texttt{Model::load\_sync} call within \texttt{py.allow\_threads} to release the GIL, and converts each tensor to the target framework format.

\subsection{ServerlessLLM Synchronous Loading}

The synchronous loading function checks the partition count from the index file and dispatches to either \texttt{Model::load\_parallel\_sync} for multi-partition checkpoints or \texttt{Model::load\_sync} for single-partition checkpoints.

\section{Python Binding Signatures}

\subsection{load\_safetensors\_sync}

\small
\begin{verbatim}
load_safetensors_sync(
    path: str,
    framework: str = "torch",
    device: str = "cpu"
) -> dict
\end{verbatim}
\normalsize

Loads a SafeTensors checkpoint synchronously. Returns a dictionary mapping tensor names to framework tensor objects.

\subsection{load\_safetensors\_mmap}

\small
\begin{verbatim}
load_safetensors_mmap(
    path: str,
    framework: str = "torch",
    device: str = "cpu"
) -> dict
\end{verbatim}
\normalsize

Opens a SafeTensors checkpoint as a memory map. Tensor access triggers on-demand page loading.

\subsection{load\_serverlessllm\_sync}

\small
\begin{verbatim}
load_serverlessllm_sync(
    path: str,
    framework: str = "torch",
    device: str = "cpu"
) -> dict
\end{verbatim}
\normalsize

Loads a ServerlessLLM-style checkpoint synchronously. Automatically selects parallel or sequential loading based on partition count.

\section{vLLM Loader Configuration}

The TensorStore loader is registered under the name \texttt{tensor\_store} and configured via \texttt{model\_loader\_extra\_config}:

\small
\begin{verbatim}
{
    "format": "safetensors" | "serverlessllm",
    "backend": "sync" | "mmap"
}
\end{verbatim}
\normalsize
